---
title: "Applied spatial data analysis with R: status and prospects"
author: "Roger Bivand"
date: "19 February 2019"
output: 
  beamer_presentation:
    theme: metropolis
    pandoc_args: [
      '--pdf-engine=xelatex'
    ]
    highlight: pygments
    includes:
      in_header: header.tex
    keep_tex: true
classoption: "aspectratio=169"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r size, echo=FALSE, results='hide'}
knitr::knit_hooks$set(mysize = function(before, options, envir) {
  if (before) 
    return(options$size)
})
knitr::opts_chunk$set(prompt=TRUE)
suppressMessages(library(extrafont))
suppressMessages(loadfonts())
```

```{r set-options, echo=FALSE, results='hide'}
options(width = 50)
```

# Introduction

## Outline

- Spatial and spatio-temporal data are characterised by structures that distinguish them from typical tabular data

- The geometric structures also have spatial reference system information, and can adhere to standards, which may ease geometrical operations

- Satellite data and numerical model output data typically have regular grid structures, but these are often domain-specific

- Computationally intensive tasks include interpolation, upsampling, focal operations, change of support and handling vector data with very detailed boundaries, as well as modelling using Bayesian inference

- A further challenge to modelling using training sets with spatial data is how to split the observations in the presence of spatial dependence

# Spatial and spatio-temporal data

## Spatial data

\begincols
\begincol{0.48\textwidth}

\endcol

\begincol{0.48\textwidth}
```{r , echo = FALSE, eval=TRUE, mysize=TRUE, size='\\tiny', cache=TRUE, warning=FALSE, results="hide"}
```

```{r, fig1, fig.show='hide', fig.height=5, fig.width=12, dev.args=list(family="Fira Sans", bg="transparent"), warning=FALSE}
```
\includegraphics[width=0.95\textwidth]{bml_bivand_feb19_files/figure-beamer/fig1-1.pdf}

\endcol
\endcols


## Vector data

\begincols
\begincol{0.48\textwidth}

\endcol

\begincol{0.48\textwidth}

\endcol
\endcols


## Raster data

\begincols
\begincol{0.48\textwidth}

\endcol

\begincol{0.48\textwidth}

\endcol
\endcols


## Spatio-temporal data

\begincols
\begincol{0.48\textwidth}

\endcol

\begincol{0.48\textwidth}

\endcol
\endcols


## Representing spatial vector data in R (**sp**)

## Representing spatial vector data in R (**sf**)

## Representing spatial raster data in R (**sp** and **raster**)

## Representing spatial raster data in R (**sf** and **stars**)


# Spatial reference systems

## Baseline WKT and PROJ4

\begincols
\begincol{0.48\textwidth}

\endcol

\begincol{0.48\textwidth}

\endcol
\endcols


## Escaping the WGS84 hub: PROJ4 6.0 and WKT2

\begincols
\begincol{0.48\textwidth}

\endcol

\begincol{0.48\textwidth}

\endcol
\endcols



# Earth observation

## Volume and frequency

\begincols
\begincol{0.48\textwidth}

\endcol

\begincol{0.48\textwidth}

\endcol
\endcols


## Array storage

\begincols
\begincol{0.48\textwidth}

\endcol

\begincol{0.48\textwidth}

\endcol
\endcols


## Cloud storage

\begincols
\begincol{0.48\textwidth}

\endcol

\begincol{0.48\textwidth}

\endcol
\endcols



# Computationally intensive tasks

## Data handling

\begincols
\begincol{0.48\textwidth}

\endcol

\begincol{0.48\textwidth}

\endcol
\endcols


## Interpolation

\begincols
\begincol{0.48\textwidth}

\endcol

\begincol{0.48\textwidth}

\endcol
\endcols


## Inference

## Inference under heterogeneity

\begincols
\begincol{0.48\textwidth}

\endcol

\begincol{0.48\textwidth}

\endcol
\endcols


## Inference under dependence

\begincols
\begincol{0.48\textwidth}

\endcol

\begincol{0.48\textwidth}

\endcol
\endcols



# Spatial dependence and subsetting data

## Heterogeneity and dependence


\begincols
\begincol{0.48\textwidth}

\endcol

\begincol{0.48\textwidth}

\endcol
\endcols


# Conclusions

## Conclusions

\begincols
\begincol{0.48\textwidth}

\endcol

\begincol{0.48\textwidth}

\endcol
\endcols



[comment]: \nobibliography{bml19}
