---
title: "Applied spatial data analysis with R: status and prospects"
author: "Roger Bivand"
date: "19 February 2019"
output: 
  beamer_presentation:
    theme: metropolis
    pandoc_args: [
      '--pdf-engine=xelatex'
    ]
    highlight: pygments
    includes:
      in_header: header.tex
    keep_tex: true
classoption: "aspectratio=169"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r size, echo=FALSE, results='hide'}
knitr::knit_hooks$set(mysize = function(before, options, envir) {
  if (before) 
    return(options$size)
})
knitr::opts_chunk$set(prompt=TRUE)
suppressMessages(library(extrafont))
suppressMessages(loadfonts())
```

```{r set-options, echo=FALSE, results='hide'}
options(width = 50)
```

# Introduction

## Outline

- Spatial and spatio-temporal data are characterised by structures that distinguish them from typical tabular data

- The geometric structures also have spatial reference system information, and can adhere to standards, which may ease geometrical operations

- Satellite data and numerical model output data typically have regular grid structures, but these are often domain-specific

- Computationally intensive tasks include interpolation, upsampling, focal operations, change of support and handling vector data with very detailed boundaries, as well as modelling using Bayesian inference

- A further challenge to modelling using training sets with spatial data is how to split the observations in the presence of spatial dependence

# Spatial and spatio-temporal data

## Spatial data

\begincols
\begincol{0.48\textwidth}

Spatial data typically combine position data in 2D (or 3D), attribute data and metadata related to the position data. Much spatial data could be called map data or GIS data. We collect and handle much more position data since global navigation satellite system (GNSS) like GPS came on stream 20 years ago, earth observation satellites have been providing data for longer. 

\endcol
\begincol{0.48\textwidth}

```{r, echo = TRUE, mysize=TRUE, size='\\tiny'}
library(osmdata)
library(sf)
bbox <- opq(bbox = 'bergen norway')
byb0 <- osmdata_sf(add_osm_feature(bbox, key = 'railway',
  value = 'light_rail'))$osm_lines
tram <- osmdata_sf(add_osm_feature(bbox, key = 'railway',
  value = 'tram'))$osm_lines
byb1 <- tram[!is.na(tram$name),]
o <- intersect(names(byb0), names(byb1))
byb <- rbind(byb0[,o], byb1[,o])
```
```{r, echo = TRUE, eval=FALSE, mysize=TRUE, size='\\tiny'}
library(mapview)
mapview(byb)
```


```{r plot1, mysize=TRUE, size='\\tiny', fig.show='hide', fig.height=4, fig.width=8, dev.args=list(family="Fira Sans")}
## library(mapview)
## mapview(byb)
## \includegraphics[width=0.95\textwidth]{bml_bivand_feb19_files/figure-beamer/plot1-1.pdf}
```

\endcol
\endcols

## Vector data

\begincols
\begincol{0.48\textwidth}

Spatial vector data is based on points, from which other geometries are constructed. The light rail tracks are 2D vector data. The points themselves are stored as double precision floating point numbers, typically without recorded measures of accuracy (GNSS provides a measure of accuracy). Here, lines are constructed from points.

\endcol
\begincol{0.48\textwidth}

```{r, echo = TRUE, mysize=TRUE, size='\\tiny'}
all(st_is(byb, "XY"))
str(st_coordinates(st_geometry(byb)[[1]]))
```

\endcol
\endcols

## Raster data

\begincols
\begincol{0.48\textwidth}

Spatial raster data is observed using rectangular (often square) cells, within which attribute data are observed. We probably do not know where within the raster cell the observed value is correct; all we know is that at the chosen resolution, this is the value representing the whole cell area.

\endcol

\begincol{0.48\textwidth}

```{r, echo = TRUE, mysize=TRUE, cache=TRUE, size='\\tiny'}
library(elevatr)
elevation <- get_elev_raster(as(byb, "Spatial"), z = 14)
```

```{r, echo = TRUE, eval=FALSE, mysize=TRUE, size='\\tiny'}
mapview(elevation, col=terrain.colors)
```

\endcol
\endcols


## Spatio-temporal data

\begincols
\begincol{0.48\textwidth}

Spatio-temporal data should be the rule but is not. Stacking things up over time to get a balanced stack is hard, and GIS representations do not handle time well. The [\textcolor{mLightBrown}{LandGIS ESA CCI land cover series}](https://landgis.opengeohub.org/#/?base=Stamen%20(OpenStreetMap)&center=60.4478,5.3352&zoom=11&opacity=80&layer=lcv_land.cover_esacci.lc.l4_c&time=2015) may also be downloaded (API coming), and displayed.

\endcol

\begincol{0.48\textwidth}

```{r, echo = TRUE, mysize=TRUE, cache=TRUE, size='\\tiny'}
library(rgdal)
fl <- "../ESACCI-LC-L4-LCCS-Map-300m-P1Y-1992_2015-v2.0.7.tif"
LUC <- readGDAL(fl, offset=c(10400,66450), region.dim=c(400, 700),
  output.dim=c(400, 700))
```

```{r, echo = TRUE, eval=FALSE, mysize=TRUE, size='\\tiny'}
mapview(LUC[, , c(1,13,24)])
```

\endcol
\endcols


## Representing spatial vector data in R (**sp**)

\begincols
\begincol{0.48\textwidth}

The **sp** package was a child of its time, using S4 formal classes, and the best compromise we then had of positional representation (not arc-node, but hard to handle holes in polygons). If we coerse `byb` to the **sp** representation, we see the formal class structure. Input/output used OGR/GDAL vector drivers in the **rgdal** package, and topological operations used GEOS in the **rgeos** package.

\endcol

\begincol{0.48\textwidth}

```{r, echo = TRUE, mysize=TRUE, size='\\tiny'}
library(sp)
str(slot(as(st_geometry(byb), "Spatial"), "lines")[[1]])
```


\endcol
\endcols

## Representing spatial vector data in R (**sf**)


\begincols
\begincol{0.48\textwidth}

The recent **sf** package bundles GDAL and GEOS (**sp** just defined the classes and methods, leaving I/O and computational geometry to other packages). **sf** used `data.frame` objects with one (or more) geometry column for vector data. The representation follows ISO 19125 (*Simple Features*), and has WKT (text) and WKB (binary) representations (used by GDAL and GEOS internally). The drivers include PostGIS and other database constructions permitting selection, and WFS for server APIs (**rgdal** does too, but requires more from the user).

\endcol

\begincol{0.48\textwidth}

```{r, echo = TRUE, mysize=TRUE, size='\\tiny'}
strwrap(st_as_text(st_geometry(byb)[[1]]))
```

\endcol
\endcols

## Representing spatial raster data in R (**sp** and **raster**)

\begincols
\begincol{0.48\textwidth}

The **raster** package S4 representation builds on the **sp** representation, using a `GridTopology` S4 object to specify the grid, and a data frame to hold the data. **raster** defines `RasterLayer`, and combinations of layers as stacks (may be different storage classes) or bricks (same storage class - array). Adding time remains an issue; **raster** could avoid reading data into memory using **rgdal** mechanisms.

\endcol

\begincol{0.48\textwidth}

```{r, echo = TRUE, mysize=TRUE, cache=TRUE, size='\\tiny'}
elevation
slot(LUC, "grid")
```


\endcol
\endcols

## Representing spatial raster data in R (**sf** and **stars**)

\begincols
\begincol{0.48\textwidth}

The new **stars** - Scalable, Spatiotemporal Tidy Arrays - package started looking at array structures and has built-in proxy data. Like **sf**, the development of **stars** has been supported by the R Consortium, and **stars** uses the infrastructure of **sf** to use GDAL for input/output and manipulation. In **sf**, the interface to the C++ GDAL library is based on **Rcpp**, which was not available when **rgdal** was written (plot from [\textcolor{mLightBrown}{colour table}](https://github.com/r-spatial/mapview/issues/208) needs current **stars** master).

\endcol

\begincol{0.48\textwidth}
```{r, echo = TRUE, mysize=TRUE, cache=TRUE, size='\\tiny'}
library(stars)
LUC_stars <- read_stars(fl, proxy=TRUE)
LUC_stars
LUC_stars1 <- st_as_stars(LUC_stars[byb,,,])
```

```{r, echo = TRUE, eval=FALSE, mysize=TRUE, size='\\tiny'}
legs <- read.csv("../ESACCI-LC-Legend.csv", header = TRUE, sep = ";")
plot(LUC_stars1, band=24, rgb = legs, main="")
```
\endcol
\endcols


# Spatial reference systems

## Baseline WKT and PROJ4

\begincols
\begincol{0.48\textwidth}

Spatial reference systems define how the geoid is viewed (prime meridian, ellipsoid, datum), and, if projected to the plane, where we are (central longitude, latitude, offsets, etc.). They also define the units - **sf** incorporates smart units handling. Projection (no datum change) and transformation are possible using PROJ and its `proj_api.h` interface directly (`rgdal::spTransform()` and `lwgeom::st_transform_proj()`), or through GDAL (`sf::st_transform()`).

\endcol

\begincol{0.48\textwidth}

```{r, echo = TRUE, mysize=TRUE, cache=TRUE, size='\\tiny'}
(WKT <- st_crs(byb))
strwrap(gsub(",", ", ", st_as_text(WKT)))
byb_utm <- st_transform(byb, crs=32632)
st_crs(byb_utm)
```

\endcol
\endcols


## Escaping the WGS84 hub: PROJ 6.0 and OGC WKT2

\begincols
\begincol{0.48\textwidth}

Changes in the legacy PROJ representation and WGS84 transformation hub have been coordinated through the [\textcolor{mLightBrown}{GDAL barn raising}](https://gdalbarn.com/) initiative. The syntax is changing to pipelines, but crucially WGS84 will often cease to be the pivot for moving between datums. A new OGC WKT is coming, and an SQLite EPSG file database will replace CSV files. SRS will begin to support 3D by default, adding time too as SRS change. 

\endcol

\begincol{0.48\textwidth}

```{r, echo = TRUE, mysize=TRUE, cache=TRUE, size='\\tiny'}
head(st_coordinates(st_geometry(byb)[[1]]), n=1)
system(paste("echo 5.333375 60.30436 |", 
  "proj +proj=pipeline +ellps=WGS84", 
  "+step +init=epsg:32632"), intern=TRUE)
head(st_coordinates(st_geometry(byb_utm)[[1]]), n=1)
```

\endcol
\endcols



# Earth observation

## Volume and frequency

\begincols
\begincol{0.48\textwidth}

As earth observation (and other remote sensor) data volume (resolution) and frequency has increased, legacy technologies of downloading (usually calibrated) scenes to multiple user systems have been struggling to keep up. Terralib is an early example of trying to keep data in databases, but maybe using WMS and other API technologies.

\endcol

\begincol{0.48\textwidth}

\includegraphics[width=0.95\textwidth]{terralib.png}

\endcol
\endcols



## Cloud storage

\begincols
\begincol{0.48\textwidth}

[\textcolor{mLightBrown}{OpenEO}](http://openeo.org/about/) proposes proof-of-concept client-server API approaches; the **elevatr** example earlier shows a specialised use. They build on data being kept in the cloud, and accessed by selection through an API. There is also some [\textcolor{mLightBrown}{client-server}](https://www.r-spatial.org/r/2018/03/23/stars3.html) code in **stars**, needing to be run in two separate, concurrent R sessions:

\endcol

\begincol{0.48\textwidth}
```{r, echo = TRUE, eval=FALSE, mysize=TRUE, cache=TRUE, size='\\tiny'}
library(plumber)
r <- plumb(system.file("plumber/server.R", package = "stars"))
r$run(port=8000)
```

```{r, echo = TRUE, eval=FALSE, mysize=TRUE, cache=TRUE, size='\\tiny'}
source(system.file("plumber/client.R", package = "stars"),,
  echo = TRUE)
plot(xx)
```

\endcol
\endcols



# Computationally intensive tasks

## Data handling

\begincols
\begincol{0.48\textwidth}

[\textcolor{mLightBrown}{city bike}](https://bergenbysykkel.no/en/open-data)

\endcol

\begincol{0.48\textwidth}

```{r, echo = TRUE, mysize=TRUE, cache=TRUE, size='\\tiny'}
bike_fls <- list.files("../bbs")
trips0 <- NULL
for (fl in bike_fls) trips0 <- rbind(trips0,
  read.csv(file.path("../bbs", fl), header=TRUE))
trips0 <- trips0[!(trips0[,9]==84),]
trips <- cbind(trips0[,c(1, 4, 2, 9)], data.frame(count=1))
from <- unique(trips0[,c(4,5,7,8)])
names(from) <- substring(names(from), 7)
to <- unique(trips0[,c(9,10,12,13)])
names(to) <- substring(names(to), 5)
stations0 <- st_as_sf(merge(from, to, all=TRUE),
  coords=c("station_longitude", "station_latitude"))
stations <- aggregate(stations0, list(stations0$station_id),
  head, n=1)
suppressWarnings(stations <- st_cast(stations, "POINT"))
st_crs(stations) <- 4326
od <- aggregate(trips[,-(1:4)], list(trips$start_station_id,
  trips$end_station_id), sum)
library(stplanr)
od_lines <- od2line(flow=od, zones=stations, zone_code="Group.1",
  origin_code="Group.1", dest_code="Group.2")
```

```{r, echo = TRUE, eval=FALSE, mysize=TRUE, cache=TRUE, size='\\tiny'}
mapview(od_lines, alpha=0.2, lwd=(od_lines$x/max(od_lines$x))*10)
```

\endcol
\endcols


## Interpolation

\begincols
\begincol{0.48\textwidth}

\endcol

\begincol{0.48\textwidth}

\endcol
\endcols


## Inference

## Inference under heterogeneity

\begincols
\begincol{0.48\textwidth}

\endcol

\begincol{0.48\textwidth}

\endcol
\endcols


## Inference under dependence

\begincols
\begincol{0.48\textwidth}

\endcol

\begincol{0.48\textwidth}

\endcol
\endcols



# Spatial dependence and subsetting data

## Heterogeneity and dependence


\begincols
\begincol{0.48\textwidth}

\endcol

\begincol{0.48\textwidth}

\endcol
\endcols


# Conclusions

## Conclusions

- 

- 

- 

- 


