---
title: "Applied spatial data analysis with R: status and prospects"
author: "Roger Bivand"
date: "19 February 2019"
output: 
  beamer_presentation:
    theme: metropolis
    pandoc_args: [
      '--pdf-engine=xelatex'
    ]
    highlight: pygments
    includes:
      in_header: header.tex
    keep_tex: true
classoption: "aspectratio=169"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r size, echo=FALSE, results='hide'}
knitr::knit_hooks$set(mysize = function(before, options, envir) {
  if (before) 
    return(options$size)
})
knitr::opts_chunk$set(prompt=TRUE)
suppressMessages(library(extrafont))
suppressMessages(loadfonts())
```

```{r set-options, echo=FALSE, results='hide'}
options(width = 50)
```

# Introduction

## Outline

- Spatial and spatio-temporal data are characterised by structures that distinguish them from typical tabular data

- The geometric structures also have spatial reference system information, and can adhere to standards, which may ease geometrical operations

- Satellite data and numerical model output data typically have regular grid structures, but these are often domain-specific

- Computationally intensive tasks include interpolation, upsampling, focal operations, change of support and handling vector data with very detailed boundaries, as well as modelling using Bayesian inference

- A further challenge to modelling using training sets with spatial data is how to split the observations in the presence of spatial dependence



[comment]: \nobibliography{bml19}
